maze-dataset

### 2025-04-08 docs/paper/paper.tex
minor paper edits, fix figure 7 success rate

\author[1,3%
%
\ensuremath\mathparagraph]{Michael Igorevich Ivanitskiy%
  \,\orcidlink{0000-0002-4213-4993}\,%
}
\author[2%
%
]{Aaron Sandoval%
  \,\orcidlink{0009-0002-8380-6140}\,%
}
\author[2,3%
%
]{Alex F. Spies%
  \,\orcidlink{0000-0002-8708-1530}\,%
}
\author[2%
%
]{Tilman Räuker%
  \,\orcidlink{0009-0009-6321-4413}\,%
}
\author[1%
%
]{Brandon Knutson%
  \,\orcidlink{0009-0004-8413-0239}\,%
}
\author[1%
%
]{Cecilia Diniz Behn%
  \,\orcidlink{0000-0002-8078-5105}\,%
}
\author[1%
%
]{Samy Wu Fung%
  \,\orcidlink{0000-0002-2926-4582}\,%
}

\affil[1]{Colorado School of Mines, Department of Applied Mathematics
and Statistics%
}
\affil[2]{Imperial College London%
}
\affil[3]{unsearch.org%
}
\affil[4]{Independent%
}
\affil[$\mathparagraph$]{Corresponding author}
\date{\vspace{-2.5ex}}

@@

In order to replicate the exact dataset distribution of
(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}), we allow for additional constraints in
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#MazeDatasetConfig.endpoint_kwargs}{\texttt{MazeDatasetConfig.endpoint\_kwargs:}}
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#EndpointKwargsType}{\texttt{EndpointKwargsType}},
such as enforcing that the start or end point be in a ``dead end'' with
only one accessible neighbor cell. However, combining this with cyclic
mazes (such as those generated with percolation), as was required for
the work in
(\protect\hyperlink{ref-knutson2024logicalextrapolation}{Knutson et al.,
2024}), can lead to an absence of valid start and end points. Placing
theoretical bounds on this success rate is difficult, as it depends on
the exact maze generation algorithm and parameters used. However, our
package provides a way to estimate the success rate of a given
configuration using a symbolic regression model trained with PySR
(\protect\hyperlink{ref-pysr}{Cranmer, 2023}). More details on this can
be found in
\href{https://understanding-search.github.io/maze-dataset/notebooks/estimate_dataset_fractions.html}{\texttt{estimate\_dataset\_fractions.ipynb}}.

Using the estimation simply requires the user to call
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDatasetConfig.success_fraction_compensate}{\texttt{cfg\_new:\ MazeDatasetConfig\ =\ cfg.success\_fraction\_compensate()}},
providing their initial \texttt{cfg} and then using the returned
\texttt{cfg\_new} in its place.

@@

However, the outputs of this function are not directly usable due to
minor divergences at the endpoints with respect to the percolation
probability \(p\). Since we know that maze success is either guaranteed
or impossible for \(p=0\) and \(p=1\), we define the
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/success_predict_math.html\#soft_step}{\texttt{soft\_step}}
function to nudge the raw output of the symbolic regression. This
function is defined with:

A shifted sigmoid \(\sigma_s\), amplitude scaling \(A\), and \(h\)
function: \[
  \sigma_s(x) = (1 + e^{-10^3 \cdot (x-0.5)})^{-1}
  \qquad A(q,a,w) = w \cdot (1 - |2q-1|^a)
\] \[
  h(q,a) = q \cdot (1 - |2q-1|^a) \cdot (1-\sigma_s(q)) + (1-(1-q) \cdot (1 - |2(1-q)-1|^a)) \cdot \sigma_s(q)
\]

@@

where \texttt{raw\_val} is the output of the symbolic regression model.
\(x_0\) is the percolation probability, while all other parameters from
\texttt{\_to\_ps\_array()} only affect \texttt{raw\_val}.

\begin{figure}
\centering
\includegraphics[width=1\textwidth,height=\textheight]{figures/ep/ep_deadends_unique.pdf}
\caption{An example of both empirical and predicted success rates as a
function of the percolation probability \(p\) for various maze sizes,
percolation with and without depth first search, and
\texttt{endpoint\_kwargs} requiring that both the start and end be in
unique dead ends.}
\end{figure}

@@

This work was partially supported by and many of the authors were
brought together by AI Safety Camp and AI Safety Support. We would like
to thank our former collaborators at AI Safety Camp and other users and
contributors to the \texttt{maze-dataset} package: Naveen Arunachalam,
Benji Berczi, Guillaume Corlouer, William Edwards, Leon Eshuijs, Chris
Mathwin, Lucia Quirke, Can Rager, Adrians Skapars, Johannes Treutlein,
and Dan Valentine.

>>>>

\author[1%
%
\ensuremath\mathparagraph]{Michael Igorevich Ivanitskiy%
  \,\orcidlink{0000-0002-4213-4993}\,%
}
\author[4%
%
]{Aaron Sandoval%
  \,\orcidlink{0009-0002-8380-6140}\,%
}
\author[2%
%
]{Alex F. Spies%
  \,\orcidlink{0000-0002-8708-1530}\,%
}
\author[3%
%
]{Tilman Räuker%
  \,\orcidlink{0009-0009-6321-4413}\,%
}
\author[1%
%
]{Brandon Knutson%
  \,\orcidlink{0009-0004-8413-0239}\,%
}
\author[1%
%
]{Cecilia Diniz Behn%
  \,\orcidlink{0000-0002-8078-5105}\,%
}
\author[1%
%
]{Samy Wu Fung%
  \,\orcidlink{0000-0002-2926-4582}\,%
}

\affil[1]{Colorado School of Mines, Department of Applied Mathematics
and Statistics%
}
\affil[2]{Imperial College London%
}
\affil[3]{UnSearch.org%
}
\affil[4]{Independent%
}
\affil[$\mathparagraph$]{Corresponding author}
\date{\vspace{-2.5ex}}

@@

In order to replicate the exact dataset distribution of
(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}), the parameter
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#MazeDatasetConfig.endpoint_kwargs}{\texttt{MazeDatasetConfig.endpoint\_kwargs:}}
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/maze_dataset_config.html\#EndpointKwargsType}{\texttt{EndpointKwargsType}}
allows for additional constraints such as enforcing that the start or
end point be in a ``dead end'' with only one accessible neighbor cell.
However, combining these constraints with cyclic mazes (such as those
generated with percolation), as was required for the work in
(\protect\hyperlink{ref-knutson2024logicalextrapolation}{Knutson et al.,
2024}), can lead to an absence of valid start and end points. Placing
theoretical bounds on this success rate is difficult, as it depends on
the exact maze generation algorithm and parameters used. To deal with
this, our package provides a way to estimate the success rate of a given
configuration using a symbolic regression model trained with PySR
(\protect\hyperlink{ref-pysr}{Cranmer, 2023}). More details on this can
be found in
\href{https://understanding-search.github.io/maze-dataset/notebooks/estimate_dataset_fractions.html}{\texttt{estimate\_dataset\_fractions.ipynb}}.
Using the estimation algorithm simply requires the user to call
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDatasetConfig.success_fraction_compensate}{\texttt{cfg\_new:\ MazeDatasetConfig\ =\ cfg.success\_fraction\_compensate()}},
providing their initial \texttt{cfg} and then using the returned
\texttt{cfg\_new} in its place.

@@

However, the outputs of this function are not directly usable due to
minor divergences at the endpoints with respect to the percolation
probability \(p\). Since we know that maze success is either guaranteed
or impossible for \(p=0\) and \(p=1\), we define the
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/success_predict_math.html\#soft_step}{\texttt{soft\_step}}
function to nudge the raw output of the symbolic regression. This
function is defined with the following components:

shifted sigmoid \(\sigma_s\), amplitude scaling \(A\), and \(h\)
function given by \[
  \sigma_s(x) = (1 + e^{-10^3 \cdot (x-0.5)})^{-1}
  \qquad A(q,a,w) = w \cdot (1 - |2q-1|^a)
\] \[
  h(q,a) = q \cdot (1 - |2q-1|^a) \cdot (1-\sigma_s(q)) + (1-(1-q) \cdot (1 - |2(1-q)-1|^a)) \cdot \sigma_s(q)
\]

@@

where \texttt{raw\_val} is the output of the symbolic regression model.
The parameter \(x_0\) is the percolation probability, while all other
parameters from \texttt{\_to\_ps\_array()} only affect
\texttt{raw\_val}.

\begin{figure}
\centering
\includegraphics[width=1\textwidth,height=\textheight]{figures/ep/ep_deadends_unique-crop.pdf}
\caption{An example of both empirical and predicted success rates as a
function of the percolation probability \(p\) for various maze sizes,
percolation with and without depth first search, and
\texttt{endpoint\_kwargs} requiring that both the start and end be in
unique dead ends.}
\end{figure}

@@

This work was partially supported by and many of the authors were
brought together by AI Safety Camp and AI Safety Support. We would like
to thank our former collaborators at AI Safety Camp and other users and
contributors to the \texttt{maze-dataset} package: Benji Berczi,
Guillaume Corlouer, William Edwards, Leon Eshuijs, Chris Mathwin, Lucia
Quirke, Can Rager, Adrians Skapars, Rusheb Shah, Johannes Treutlein, and
Dan Valentine.
